---
title: "exercise-10"
format: html
editor: visual
---

#Challenge 1

```{r}
library(tidyverse)
library(skimr)

install.packages("skimr")

f <- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/AVONETdataset1.csv"

d <- read.csv(f, header = TRUE)
d <- as_tibble(d)
d <- d |> select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle,Range.Size)

skim(d)

#convert migration from numeric to as factor

d <- d |> mutate(Migration = as.factor(Migration))

# make boxplots 
library(ggplot2)
d <- drop_na(d, Trophic.Level, Migration)

#By Trophic.Level
ggplot(d, aes(x = Trophic.Level, y = log(Mass))) +
  geom_boxplot() +
  labs(title = "Boxplot of log(Mass) by Trophic Level",
       x = "Trophic Level",
       y = "log(Mass)") +
  theme_minimal()

#By Migration
ggplot(d, aes(x = Migration, y = log(Mass))) +
  geom_boxplot() +
  labs(title = "Boxplot of log(Mass) by Migration",
       x = "Migration Type",
       y = "log(Mass)") +
  theme_minimal()

```
#Step 1: Making boxplots for the variables

```{r}
ggplot( data= d |> drop_na(Trophic.Level), aes(x= Trophic.Level, y= log(Mass))) +
          geom_boxplot() + geom_jitter()  


d <- d |> mutate(Migration = as.factor(Migration))

ggplot( data = d |> drop_na(Migration), aes(x= Migration, y =       log(Mass))) + geom_boxplot() + geom_jitter()

```


#Step 2: Run linear models

```{r}
 m1 <- lm(log(Mass) ~ Trophic.Level, data = d)
 summary(m1)

 m2 <- lm(log(Mass) ~ Migration, data = d)
 summary(m2)
 
 # setting Migration2 as the reference level
d$Migration <- relevel(d$Migration, ref = "2")

m3 <- lm(log(Mass) ~ Migration, data = d)
summary(m3)

 
 #Answer: Migration 1 and 3 are significantly different from the reference level (Migration2). The negative coefficients indicate that the log(Mass) for these migration categories is lower than that of Migration2.
```

The `echo: false` option disables the printing of code (only output is displayed).

#Step 3: Conducting a TukeyHSD Test
```{r}

m3 <- lm(log(Mass) ~ Migration, data = d)
pairwise.t.test(log(d$Mass), d$Migration, p.adj = "bonferroni")

m3 <- aov(log(Mass) ~ Migration, data = d)
posthoc <- TukeyHSD(m3, which = "Migration", ordered = TRUE, conf.level = 0.95)
posthoc

plot(posthoc, xlim= c(0,1))

# There is a significant difference between migration 1 and 2


```


#Step 4: Permutation
```{r}

library(infer)

d <- d|> mutate(logMass = log(Mass))
perm.F <- d |> specify(logMass ~ Trophic.Level) |>
  hypothesise(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate( stat = "F")

original.F <- aov(log(Mass) ~ Trophic.Level, data = d) |>
  broom:: tidy() |>
  filter(term== "Trophic.Level")

original.F

visualise(perm.F)
visualise(perm.F) + shade_p_value(obs_stat = original.F$statistic, direction = "greater")
```


#Challenge 2 : Data Wrangling, One and Two- Factor ANOVA

#Step 1: Creating new variables in the dataset

```{r}
beakl <- lm(log(Beak.Length_Culmen) ~ log(Mass), data = d)
tarsusl<- lm(log(Tarsus.Length) ~ log(Mass), data=d)

summary(beakl)
summary(tarsusl)

d <- d |> mutate(rel_beak_length =  beakl$residuals)
d <- d |> mutate(rel_tarsus_length = tarsusl$residuals)

```


#Step 2: Boxplot and violin plot
```{r}
ggplot(data = d, aes(x = Primary.Lifestyle, y = rel_tarsus_length)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data = d |> drop_na(Trophic.Niche), aes(x = Trophic.Niche, y = rel_beak_length)) +
  geom_violin() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data = d |> drop_na(Trophic.Niche), aes(x = Trophic.Niche, y = rel_beak_length)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#Step 3: Run ANOVA Analyses
```{r}

migration <- d |> drop_na(Migration)
hist(migration$Range.Size)

migration <- migration |>mutate(logRS = log(Range.Size))
hist(migration$logRS)

run_a <- lm(data = migration, logRS ~ Migration)
summary(run_a)

# Answer: The range is associated with the migration group because the F statistic and p-values are significant enough to reject the null. The 8.69% of the variance is explained by the regression model. 

#Answer2: The new reference level is now migration category 2. MIgrations 1 and 3 differ from the rference level as migration 1 is lower and migration 3 is higher tham 2 respectively. 


migration <- migration |> mutate(Migration= relevel(as.factor(Migration), ref = "3"))

run_a2 <- lm(data= migration, logRS ~ Migration)
summary(run_a2)

m3 <- aov(logRS ~ Migration, data = migration)
summary(m3)


mighoc<- TukeyHSD(m3, which = "Migration", ordered = TRUE, conf.level = 0.95)

mighoc
plot(mighoc, xlim = c(0,5))

#Answer: Migration 1 and 2 have lower mean logRS than Migration 3. Importantly there is a significant difference between 1 and 3 


```

#Step 4: Winnowinf for Passeriformes
```{r}
passeriformes <- d |> filter(Order1 =="Passeriformes")
ggplot(passeriformes, aes(x= Primary.Lifestyle, y= rel_beak_length, fill= Primary.Lifestyle)) + geom_boxplot() + 
  labs(title = "Boxplot of Relative Beak length by Primary Lifestyle")


ggplot(passeriformes, aes(x= Trophic.Level, y= rel_beak_length, fill = Trophic.Level)) + geom_boxplot() + 
  labs(title = "Boxplot of Relative Beak length by Trophic Level")


ggplot(passeriformes, aes(x= interaction(Primary.Lifestyle, Trophic.Level), y= rel_beak_length)) + geom_boxplot() + 
  labs(title = "Boxplot of Relative Beak length by Primary LIfestyle and Trophic Level") + theme(axis.text.x= element_text(angle = 45, hjust = 1))


p1 <- lm(data = passeriformes, rel_beak_length ~ Primary.Lifestyle)
p2 <- lm(data = passeriformes, rel_beak_length ~ Trophic.Level)

summary(p1)
summary(p2)

#Answer: The Primary Lifestyle model shows significant effects on relative beak length for all categories, with Insessorial and Terrestrial birds having the largest beaks. The model explains 5.5% of the variance. The Trophic Level model reveals that herbivores have smaller beaks, while omnivores have slightly larger beaks, explaining 5% of the variance. Both models are statistically significant, but they explain a small portion of the variance in beak length.

```
#Step 5: Two factor model on relative beak length
```{r}

two_f_anova <- lm(rel_beak_length ~ Primary.Lifestyle + Trophic.Level, data = passeriformes)
summary(two_f_anova)

two_f_anova <- aov(rel_beak_length ~ Primary.Lifestyle + Trophic.Level, data = passeriformes)
summary(two_f_anova)

#Answer: The two-factor ANOVA shows that both Primary.Lifestyle (F = 136.8, p < 2e-16) and Trophic.Level (F = 201.3, p < 2e-16) significantly affect relative beak length in passeriforms, indicating both factors are strongly related to beak size variations.

```

#Step 6: Tw0-way Model
```{r}

m8 <- lm(rel_beak_length ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = passeriformes)
summary(m8)

m8<- aov( rel_beak_length ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = passeriformes)
summary(m8)

m9<- aov( rel_beak_length ~ Primary.Lifestyle* Trophic.Level, data = passeriformes)
summary(m9)

#Answer: Primary.Lifestyle and Trophic.Level independently influence relative beak length, but their interaction also plays a significant role in shaping this relationship.

```
#Step 7: Interactive plot
```{r}

interaction.plot(x.factor = passeriformes$Primary.Lifestyle, trace.factor = passeriformes$Trophic.Level, response = passeriformes$rel_beak_length, fun = base::mean, xlab = "Primary Lifestyle", ylab = "Mean Relative Beak Length", trace.label = "Trophic Level") 

```

#Step 8: Comparing large and small sd
```{r}


stats <- passeriformes |> 
  group_by(Trophic.Level) |> 
  summarise(
    mean_rel_beak_length = mean(rel_beak_length), 
    sd_rel_beak_length = sd(rel_beak_length)
  )

sd_ratio <- max(stats$sd_rel_beak_length) / min(stats$sd_rel_beak_length)
print(sd_ratio)

par(mfrow = c(1,2))

trophic_levels <- c("Carnivore", "Herbivore", "Omnivore")

for (level in trophic_levels) {
  data <- passeriformes$rel_beak_length[passeriformes$Trophic.Level == level]
  
  hist(data, main = level, xlab = "Relative beak length")
  
  qqnorm(data, main = paste("QQ Plot -", level))
  qqline(data) 
}

par(mfrow = c(1,1))
```




